---
title: "Babynames_Family"
author: "Nathan W. Van Bibber"
date: "5/6/2020"
output: html_document
---

# Family Baby Names

First let's load some packages. 

```{r, results='hide'}
library(babynames)
library(dplyr)
library(ggplot2)
```

Now, let's take a glimpse at the babynames package.  For more on the `glimpse()` function type `?tibble::glimpse` since dplyr imports the function from the tibble package.
```{r}
glimpse(babynames)

unique(babynames$year)
```

## Making some plots

Now, I'll make a data frame for each of our names and plot them individually.

```{r}
babynames_lily <- babynames %>% 
    filter(name == "Lily") %>% 
    group_by(name, year) %>% 
    summarise(n_total = sum(n))

babynames_nathan <- babynames %>% 
    filter(name == "Nathan") %>% 
    group_by(name, year) %>% 
    summarise(n_total = sum(n))

babynames_cassie <- babynames %>% 
    filter(name == "Cassie") %>% 
    group_by(name, year) %>% 
    summarise(n_total = sum(n))

ggplot(babynames_lily, aes(x = year, y = n_total)) +
    geom_line() + 
    ggtitle("Popularity of the name Lily over time")

ggplot(babynames_cassie, aes(x = year, y = n_total)) +
    geom_line() + 
    ggtitle("Popularity of the name Cassie over time")

ggplot(babynames_nathan, aes(x = year, y = n_total)) +
    geom_line() + 
    ggtitle("Popularity of the name Nathan over time")
```


## Groupings

Now I'll put all of us on the same plot.

```{r}
babynames_family <- babynames %>% 
    filter(name %in% c("Cassie", "Lily", "Nathan")) %>% 
    group_by(name, year) %>% 
    summarise(n_total = sum(n))

#babynames_family

ggplot(babynames_family, aes(x = year, y = n_total, color = name)) + 
    geom_line() + 
    ggtitle("Popularity of our Family names over time.")

```


```{r}
babynames_grandparents <- babynames %>% 
    filter(name %in% c("Sharon", "Peggy", "Walter", "Connie", "Alice")) %>% 
    group_by(name, year) %>% 
    summarise(n_total = sum(n))


ggplot(babynames_grandparents, aes(x = year, y = n_total, color = name)) + 
    geom_line() + 
    ggtitle("Popularity of names over time.")

```


Now let me compare my siblings and myself.

```{r}
babynames_siblings <- babynames %>% 
    filter(name %in% c("Nathan", "Shelley", "Rachel")) %>% 
    group_by(name, year) %>% 
    summarise(n_total = sum(n))


ggplot(babynames_siblings, aes(x = year, y = n_total, color = name)) + 
    geom_line() + 
    ggtitle("Popularity of names over time.")


```


## All together now.

now everybody at once.

```{r}
babynames_all <- babynames %>% 
    filter(name %in% c("Cassie", "Lily", "Nathan", "Sharon", "Peggy", "Walter", "Connie", "Alice", "Shelley", "Rachel")) %>% 
    group_by(name, year) %>% 
    summarise(n_total = sum(n))


ggplot(babynames_all, aes(x = year, y = n_total, color = name)) + 
    geom_line() + 
    ggtitle("Popularity of names over time.")


```


## Most Unique

```{r}
babynames_unique <- babynames %>% 
    filter(name %in% c("Cassie", "Shelley")) %>% 
    group_by(name, year) %>% 
    summarise(n_total = sum(n))


ggplot(babynames_unique, aes(x = year, y = n_total, color = name)) + 
    geom_line() + 
    ggtitle("Popularity of names over time.")
```

## Other Inquireries

What were the most popular male and female names in our birth years?

```{r}
## Babynames

mostPopular_byYear <- babynames %>% 
    group_by(year, sex) %>% 
    top_n(1, n)

filter(mostPopular_byYear, year %in% c(1949,1957,1986:1988,1992,2017)) %>% 
    arrange(sex)
 
```

